---
title: "MAnorm2 for Normalizing and Comparing ATAC wD15_bD15 Samples Genrich peaks"
author: "Mohamed Abdelhalim"
date: '2022-10-11'
output:
  html_document: default
  keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = "png",
                      dpi = 300,
                      echo = FALSE,
                      cache = TRUE)
```

## Input Data

For employing the machinery implemented in MAnorm2, you need to prepare a table that profiles the ATAC signal in each of a list of genomic intervals for each of a set of ChIP-seq samples.

```{r dataset}
library(MAnorm2)
library(readr)
setwd("~/ATAC/MAnorm2/wD15_bD15/Genrich/")
bD15_wD15_profile_bins <- read_delim("bD15_wD15_Gen_profile_bins.xls", delim = "\t", escape_double = FALSE,trim_ws = TRUE)
head(bD15_wD15_profile_bins)

```

#### Normalization

```{r cmpBioReps}
# Perform within-group normalization.
norm <- normalize(bD15_wD15_profile_bins, count = 4:5, occupancy = 8:9)
norm <- normalize(norm, count = 6:7, occupancy = 10:11)

# Construct a bioCond for each group of samples.

conds <- list(bD15 = bioCond(norm[4:5], norm[8:9], name = "bD15"),
              wD15 = bioCond(norm[6:7], norm[10:11], name = "wD15"))

# Perform between-group normalization.
# Restrict common peak regions to autosomes only when the two groups
# being compared are associated with different genders.
conds <- normBioCond(conds)

# Fit a mean-variance curve.
# If the following function call raises an error,
# set init.coef = c(0.1, 10) or try method = "local".
init.coef = c(0.1, 10)
conds <- fitMeanVarCurve(conds, method = "local", occupy.only = TRUE)

# Perform differential tests.
res <- diffTest(conds[[1]], conds[[2]])
head(res)
```

In this data frame, Mval could be interpreted as log2 fold change; pval assesses the statistical significance of each test; padj refers to adjusted p-value with the "BH" method. Mval Difference in mean signal intensity between the two conditions. An Mval of 1 indicates a twofold change in normalized read count.

#### MA plots Before and After Normalization

```{r MAplotDefault, fig.show = "hold", fig.height = 4.7, fig.width = 4.7}
# Before normalization.
raw <- log(bD15_wD15_profile_bins[4:5] + 0.5, base = 2)
MAplot(raw[[1]], raw[[2]],norm[[8]],norm[[9]], ylim = c(-2, 2),
       main = "bD15_wD15 Before normalization")
abline(h = 0, lwd = 2, lty = 5)

# After normalization.
MAplot(norm[[4]], norm[[5]], norm[[8]], norm[[9]], ylim = c(-2, 2),
       main = "bD15_wD15  After normalization")
abline(h = 0, lwd = 2, lty = 5)
```

```{r MAplotBioCond2}
MAplot(conds[[1]], conds[[2]], ylim = c(-5, 5), main = "bD15 vs. wD15")
abline(h = 0, lwd = 2, lty = 5)
```

```{r summaryMVC}
summary(conds$bD15)
summary(conds$wD15)
```

#### You can visualize the overall test results by

```{r MAplotBioCond}
MAplot(res, padj = 0.01 , ylim = c(-5, 5))
abline(h = 0, lwd = 2, lty = 5, col = "green3")
```

#calculate fold change

```{r fold change}
FC <- log2((rowMeans(bD15_wD15_profile_bins[,4:5])+0.1)/(rowMeans(bD15_wD15_profile_bins[,6:7])+0.1))
results <- cbind(bD15_wD15_profile_bins[,1:3],res,FC)
head(results)
```

#Saving the Results

```{r Saving the Results}
names <- seq(dim(results)[1])
names <- paste0("peak_",names)
strand <- rep("*",dim(results)[1])
formatedresults <- cbind(results[,1:3],names,results[,3]-results[,2],strand,results$pval,results$padj,results$FC)

tab <- formatedresults[order(results$padj),]

write.table(x = tab, 
            file = "~/ATAC/MAnorm2/wD15_bD15/Genrich/Genrich_MAnorm2_pooledpeaks.bed", 
            quote = F , 
            col.names = F , 
            row.names = F , 
            sep = "\t" )

head(tab)
```

#Saving the Results for each condtion

```{r Saving the Results for each condtion}
library (dplyr)
names <- seq(dim(results)[1])
names <- paste0("peak_",names)
strand <- rep("*",dim(results)[1])
formatedresults <- cbind(results[,1:3],names,results$Mval,strand,results$pval,results$padj   ,results$FC)

# for wD15

formatedresults_wD15 <-  formatedresults  %>% filter (`results$Mval` > 0& `results$padj` < 0.01 , na.rm = TRUE)


write.table(x = formatedresults_wD15, 
            file = "~/ATAC/MAnorm2/wD15_bD15/Genrich/Genrich_MAnorm2_wD15_peaks.bed", 
            quote = F , 
            col.names = F , 
            row.names = F , 
           sep = "\t" )
head(formatedresults_wD15)

hist(formatedresults_wD15$`results$Mval`, breaks = 100, col = "red"  )
summary(formatedresults_wD15$`results$Mval`)

# for bD15

formatedresults_bD15 <-  formatedresults  %>% filter (`results$Mval` < 0 & `results$padj` < 0.01 , na.rm = TRUE)


write.table(x = formatedresults_bD15, 
            file = "~/ATAC/MAnorm2/wD15_bD15/Genrich/Genrich_MAnorm2_bD15_peaks.bed", 
            quote = F , 
            col.names = F , 
            row.names = F , 
            sep = "\t" )
head(formatedresults_bD15)

hist(abs(formatedresults_bD15$`results$Mval`), breaks = 100, col = "red")


summary(abs(formatedresults_bD15$`results$Mval`))


```

```{r Volcano plots , fig.show = "hold", fig.height = 10, fig.width = 10}

#if (!requireNamespace('BiocManager', quietly = TRUE))
    #install.packages('BiocManager')

 # BiocManager::install('EnhancedVolcano')
  
library(EnhancedVolcano)
  
EnhancedVolcano(formatedresults,
    lab = rownames(formatedresults),  
    labSize = 0 ,
    x = 'results$Mval' ,
    y = 'results$padj' ,
    title = 'wD15_bD15',
    pCutoff = 10e-2,
    FCcutoff = 0.5,
    colAlpha = 0.9,
    labCol = 'red' ,
    pointSize = 3.0)

```
#Saving the Results for each condtion after filtering with 1q 

```{r Saving the Results for each condtion after filter}
# for wD15

formatedresults_wD15_filter <-  formatedresults  %>% filter (`results$Mval` > 0.5396 & `results$padj` < 0.01 , na.rm = TRUE)


write.table(x = formatedresults_wD15_filter, 
            file = "~/ATAC/MAnorm2/wD15_bD15/Genrich/Genrich_MAnorm2_wD15_peaks_filter.bed", 
            quote = F , 
            col.names = F , 
            row.names = F , 
           sep = "\t" )

hist(formatedresults_wD15_filter$`results$Mval`, breaks = 100, col = "red"  )

# for bD15

formatedresults_bD15_filter <-  formatedresults  %>% filter (`results$Mval` < -0.6401 & `results$padj` < 0.01 , na.rm = TRUE)


write.table(x = formatedresults_bD15_filter, 
            file = "~/ATAC/MAnorm2/wD15_bD15/Genrich/Genrich_MAnorm2_bD15_peaks_filter.bed", 
            quote = F , 
            col.names = F , 
            row.names = F , 
            sep = "\t" )

hist(abs(formatedresults_bD15_filter$`results$Mval`), breaks = 100, col = "red")

```

### Peak Annotation
```{r Setting up Peak Annotation, error= TRUE , echo = FALSE}

#BiocManager::install("ChIPseeker")
#BiocManager::install("TxDb.Hsapiens.UCSC.hg38.knownGene")
#BiocManager::install("EnsDb.Hsapiens.v86")
#BiocManager::install("AnnotationDbi")
#BiocManager::install("clusterProfiler")
#BiocManager::install("org.Hs.eg.db")

# Load libraries

library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(EnsDb.Hsapiens.v86)
library(clusterProfiler)
library(AnnotationDbi)
library(org.Hs.eg.db)

```

## Loading data

```{r Setting up}
setwd("~/ATAC/MAnorm2/wD15_bD15/Genrich/Annotation/")
samplefiles <- list.files("./Genrich/Annotation", pattern= ".bed", full.names=T)
samplefiles <- as.list(samplefiles)
names(samplefiles) <- c("bD15", "Wd15","back_ground")

# Next, we need to assign annotation databases generated from UCSC to a variable

txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

#Let’s start by retrieving annotations for our peaks :

peakAnnoList <- lapply(samplefiles, annotatePeak, TxDb=txdb, 
                       tssRegion=c(-2000, 2000), verbose=FALSE)

# take a look at what is stored in peakAnnoList, you will see that the peak annotations have been summarized for you based on genomic features:

peakAnnoList

```
# Barchart of genomic feature representation

```{r Barchart of genomic feature representation}
#install.packages("ggupset")
#install.packages("ggimage")
library(ggupset)
library(ggimage)
plotAnnoBar(peakAnnoList)
plotAnnoPie(peakAnnoList[["bD15"]],main="bD15")
plotAnnoPie(peakAnnoList[["Wd15"]],main="Wb15")
upsetplot(peakAnnoList[["bD15"]],vennpie=TRUE)+ggtitle("bD15")
upsetplot(peakAnnoList[["Wd15"]],vennpie=TRUE)+ggtitle("Wd15")
```

#Distribution of TF-binding loci relative to TSS

```{r Distribution of TF-binding loci relative to TSS}
plotDistToTSS(peakAnnoList, title="Distribution of transcription factor-binding loci \n relative to TSS")


```
# Get the annotation data 
```{r Get the annotation data}
wD15_annot <- data.frame(peakAnnoList[["Wd15"]]@anno)
bD15_annot <- data.frame(peakAnnoList[["bD15"]]@anno)
back_ground_annot<- data.frame(peakAnnoList[["back_ground"]]@anno)
# Get the entrez IDs
entrez_wD15 <- wD15_annot$geneId
entrez_bD15 <- bD15_annot$geneId
entrez_back_ground <- back_ground_annot$geneId
# Return the gene symbol for the set of Entrez IDs
annotations_edb_wD15 <- AnnotationDbi::select(EnsDb.Hsapiens.v86,
                                         keys = entrez_wD15,
                                         columns = c("GENENAME"),
                                         keytype = "ENTREZID")

annotations_edb_bD15 <- AnnotationDbi::select(EnsDb.Hsapiens.v86,
                                         keys = entrez_bD15,
                                         columns = c("GENENAME"),
                                         keytype = "ENTREZID")

annotations_edb_back_ground <- AnnotationDbi::select(EnsDb.Hsapiens.v86,
                                         keys = entrez_back_ground,
                                         columns = c("GENENAME"),
                                         keytype = "ENTREZID")



# Change IDs to character type to merge
annotations_edb_wD15$ENTREZID <- as.character(annotations_edb_wD15$ENTREZID)
annotations_edb_bD15$ENTREZID <- as.character(annotations_edb_bD15$ENTREZID)
annotations_edb_back_ground$ENTREZID <-  as.character(annotations_edb_back_ground$ENTREZID)



# Write to file
library(dplyr)
wD15_annot %>% 
  left_join(annotations_edb_wD15, by=c("geneId"="ENTREZID")) %>% 
  write.table(file="./Genrich/Annotation/wD15_peak_annotation.txt", sep="\t", quote=F, row.names=F)
bD15_annot %>% 
  left_join(annotations_edb_bD15, by=c("geneId"="ENTREZID")) %>% 
  write.table(file="./Genrich/Annotation/bD15_peak_annotation.txt", sep="\t", quote=F, row.names=F)

write.table(annotations_edb_wD15,file="./Genrich/Annotation/wD15_peak_annotation_gene_list.txt", sep="\t", quote=F, row.names=F)
write.table(annotations_edb_bD15,file="./Genrich/Annotation/bD15_peak_annotation_gene_list.txt", sep="\t", quote=F, row.names=F)
write.table(annotations_edb_back_ground,file="./Genrich/Annotation/back_ground_peak_annotation_gene_list.txt", sep="\t", quote=F, row.names=F)




```

# Functional enrichment: R-based tools

```{r Functional enrichment, error= TRUE , echo = FALSE , fig.show = "hold", fig.height = 15, fig.width = 15}
ego_wD15 <- enrichGO(gene = entrez_wD15, 
                    keyType = "ENTREZID", 
                    OrgDb = org.Hs.eg.db,
                    universe = entrez_back_ground,
                    ont = "BP", 
                    pAdjustMethod = "BH", 
                    qvalueCutoff = 0.05, 
                    readable = TRUE)
ego_bD15 <- enrichGO(gene = entrez_bD15, 
                    keyType = "ENTREZID", 
                    OrgDb = org.Hs.eg.db,
                    universe = entrez_back_ground,
                    ont = "BP", 
                    pAdjustMethod = "BH", 
                    qvalueCutoff = 0.05, 
                    readable = TRUE)
dotplot(ego_wD15, showCategory=30) + ggtitle("wD15")
dotplot(ego_bD15, showCategory=30) + ggtitle("bD15")
```

# Dotplot visualization
to get Dotplot visualization
```{r Dotplot visualization, error=, include=FALSE}
jpeg("./Genrich/Annotation/dotplot-ego_wD15.jpg", width = 1000, height = 1200,res = 100)
dotplot(ego_wD15, showCategory=30) + ggtitle("wD15")
dev.off()
jpeg("./Genrich/Annotation/dotplot-ego_bD15.jpg", width = 1000, height = 1200,res = 100)
dotplot(ego_bD15, showCategory=30) + ggtitle("bD15")
dev.off()
jpeg("./Genrich/Annotation/barplot-ego_wD15.jpg", width = 1000, height = 1200,res = 100)
mutate(ego_wD15, qscore = -log(p.adjust, base=10)) %>% 
    barplot(x="qscore", showCategory=30)+ ggtitle("wD15")
dev.off()
jpeg("./Genrich/Annotation/barplot-ego_bD15.jpg", width = 1000, height = 1200,res = 100)
mutate(ego_bD15, qscore = -log(p.adjust, base=10)) %>% 
    barplot(x="qscore", showCategory=30)+ ggtitle("bD15")
dev.off()
```
# Output results from GO analysis to a table

```{r Output}
cluster_summary_wD15 <- data.frame(ego_wD15)
cluster_summary_bD15 <- data.frame(ego_bD15)
write.csv(cluster_summary_wD15,"./Genrich/Annotation/clusterProfiler_wD15.csv")
write.csv(cluster_summary_bD15,"./Genrich/Annotation/clusterProfiler_bD15.csv")
```

# KEGG pathway enrichment

```{r KEGG pathway, echo = FALSE, error = TRUE, fig.show = "hold", fig.height = 15, fig.width = 15}
ekegg_bD15 <- enrichKEGG(gene = entrez_bD15,
                 organism = 'hsa',
                 universe = entrez_back_ground,
                 pvalueCutoff = 0.05)
ekegg_wD15 <- enrichKEGG(gene = entrez_wD15,
                 organism = 'hsa',
                 universe = entrez_back_ground,
                 pvalueCutoff = 0.05)
jpeg("./Genrich/Annotation/dotplot-ekegg_bD15.jpg", width = 600, height = 600,res = 100)
dotplot(ekegg_bD15,title="bD15", showCategory=30,font.size = 12)
dev.off() 
jpeg("./Genrich/Annotation/dotplot-ekegg_wD15.jpg", width = 1000, height = 1200, res = 100)
dotplot(ekegg_wD15,title="wD15", showCategory=30,font.size = 12)
dev.off()
dotplot(ekegg_bD15,title="bD15", showCategory=30,font.size = 12)
dotplot(ekegg_wD15,title="wD15", showCategory=30,font.size = 12)
cluster_summary_wD15 <- data.frame(ekegg_wD15)
cluster_summary_bD15 <- data.frame(ekegg_bD15)
write.csv(cluster_summary_wD15,"./Genrich/Annotation/clusterProfiler-kegg_wD15.csv")
write.csv(cluster_summary_bD15,"./Genrich/Annotation/clusterProfiler_kegg_bD15.csv")

```
# Comparing enrichment across samples

```{r Comparing enrichment across samples , echo = FALSE, error = TRUE, fig.show = "hold", fig.height = 10, fig.width = 10}
# Create a list with genes from each sample
genes = lapply( peakAnnoList , function(i) as.data.frame(i)$geneId)
genes2 = genes[c(1,2)]
# Run KEGG analysis
compKEGG <- compareCluster(geneCluster = genes2, 
                         fun = "enrichKEGG",
                         organism = "human",
                         pvalueCutoff  = 0.05, 
                         pAdjustMethod = "BH")
jpeg("./Genrich/Annotation/dotplot-compKEGG.jpg", width = 1000, height = 1200, res = 100)
dotplot(compKEGG, showCategory = 20, title = "KEGG Pathway Enrichment Analysis",font.size = 10)
dev.off()
dotplot(compKEGG, showCategory = 20, title = "KEGG Pathway Enrichment Analysis",font.size = 10)
cluster_summary_wD15_bD15 <- data.frame(compKEGG)
write.csv(cluster_summary_wD15_bD15,"./Genrich/Annotation/clusterProfiler_compKEGG_wD15_bD15.csv")
```
