---
title: "Gene_expression_in_DARs"
output: html_document
date: "2024-12-11"
---

```{r message=FALSE, warning=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(ggpubr)
library(ggbeeswarm)
```


```{r theme}
time_cols = c("#37B7D2","#80D0D8","#B9D8D7","#DDDDDC","#FAE8D7","#FCB74B","#EF8E21")
theme_hp = theme_classic(base_size=15) + 
		  theme(axis.line = element_line(linewidth = 0.25),
		        axis.ticks = element_line(linewidth=0.25),
		        strip.background = element_blank(), # remove boxes around facet labels
		        )
theme_set(theme_hp)
knitr::opts_chunk$set(dev=c('png','pdf'))
```


## Load timecourse RNA-seq
caveat: this was normalised with the conversion and donor7 samples
can do a renomalisation for this paper alone for 'completeness'

```{r}
load("D:/UiO/Projects/rnaseq/00_summary_files/all_fpkms_for_plotting.RData")
subjects = data.frame(donor=c("D06G","D07G","D08A","D09A","D13A","D15LH"),
                      subject=c("S3","S4","S2","S1","S6","S5"))
subjects = subjects[order(subjects$subject),]
long$donor = factor(long$donor, levels= subjects$donor)
head(long)
```

```{r}
d13_tc = filter(long, donor=="D13A" & time %in% paste0("day", c(0,1,3,15)) &
                rep %in% c(1,2,3)) #drop extra replicates
d13_tc$group = factor(d13_tc$group, levels = c("day15_white","day3_white","day1_white",
                                               "day0_ASC","day1_beige","day3_beige","day15_beige"))
table(d13_tc$group)
table(d13_tc$rep, d13_tc$group)
```
# Load DAR gene lists

```{r}
library(here); i_am("gene_expr_in_DARs_2F.Rmd")
wDar_genes = read.delim("DAR_D15w_promoter_genelist.txt")$gene.name
head(wDar_genes); length(wDar_genes)
bDAR_genes = read.delim("DAR_D15b_promoter_genelist.txt")$gene.name
head(bDAR_genes); length(bDAR_genes)

summary(bDAR_genes %in% wDar_genes) #no overlaps
total_DARg = c(bDAR_genes, wDar_genes)
summary(total_DARg %in% long$gene_name) # we miss 87 genes with 0 expression
missing = total_DARg[!total_DARg %in% long$gene_name]
missing
```
### Filter and Fill 0s

```{r}
dar_fpkm = filter(d13_tc, gene_name %in% total_DARg) 
dar_fpkm$gene_name = factor(dar_fpkm$gene_name,levels = c(unique(dar_fpkm$gene_name), missing))
table(dar_fpkm$group)

dar_fpkm = complete(dar_fpkm, gene_name, group, rep, fill= list(fpkm=0)) 

dar_fpkm =  mutate(dar_fpkm, dar_type = if_else(gene_name %in% wDar_genes, "white", 
                            if_else(gene_name %in% bDAR_genes, "beige",NA)),
         name = paste0(group,"_rep", rep)) 
dar_fpkm$dar_type = factor(dar_fpkm$dar_type, levels=c("white","beige"))

table(dar_fpkm$group, useNA = "ifany")
table(dar_fpkm$name, useNA = "ifany")
```

```{r}
tail(dar_fpkm)
table(dar_fpkm$dar_type, useNA = "ifany")
```
### Calculate averages

And technically we want to plot the spread as an average across replicates

```{r}
summary= group_by(dar_fpkm, group, gene_name, dar_type) %>% summarise(mean_expr = mean(fpkm)) #average across reps
head(summary)
write.table(summary, here("mean_expr_for_genes_in_dars.tsv"), sep="\t", quote=FALSE, row.names = F)
rep_avg =  group_by(dar_fpkm, group, rep, dar_type) %>% summarise(mean_expr = mean(fpkm),
                                                                  log_mean_expr = log2(mean(fpkm+1))) #average across genes
head(rep_avg)
write.table(summary, here("mean_expr_per_replicate.tsv"), sep="\t", quote=FALSE, row.names = F)
```
Testing log2 scaled RNA expression (FPKM) . 

## Two way ANOVA
```{r}
summary(aov(log_mean_expr ~ dar_type + group, data=rep_avg ))
```

## T-test for individual comparisons
```{r}
rna_stats = compare_means(log_mean_expr ~ group, data= rep_avg, group.by="dar_type", method = "t.test", 
                          p.adjust.method = "none")
comparisons = c("day0_ASC;day1_beige","day1_beige;day3_beige","day15_white;day0_ASC",
                "day3_beige;day15_beige","day0_ASC;day15_beige","day15_white;day15_beige")
#Adjust only on comparisons of interest
rna_stats = mutate(rna_stats, comp = paste(group1, group2, sep=";")) %>% 
  filter(comp %in% comparisons) %>%  mutate(p.adj = p.adjust(p, method="holm"), p.adj.format = format.pval(p.adj, 3))
rna_stats[c(1,3:6,11)]
write.table(rna_stats, here("t.test_mean_between_reps_result.tsv"), sep="\t", quote=FALSE, row.names = F)
```

## Figures

Boxplot shows variation in expression between genes in each category (DAR type = white or beige).  
raw plots list unadjusted pvalues; add adj pvals in illustrator. 

```{r}
ggplot(summary, aes(x=group, y=log2(mean_expr+1), fill=group)) +geom_boxplot(outlier.shape=NA, linewidth=0.25) + 
  scale_fill_manual(values=time_cols) + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  facet_wrap(~dar_type) + 
  geom_beeswarm(data=rep_avg, aes(y=log2(mean_expr+1)), cex = 3) +
  coord_cartesian(y=c(0,12)) +
  stat_compare_means(data=rep_avg, comparisons = list(c("day0_ASC","day1_beige"),
                                        c("day1_beige","day3_beige"),
                                        c("day15_white","day0_ASC"),
                                        c("day3_beige","day15_beige"),
                                        c("day15_beige","day0_ASC"), 
                                        c("day15_white","day15_beige")),
                     method="t.test", label="p.signif", label.y = seq(from=7, to=11, length.out=6),
                     tip.length =0.01)


ggsave(here("log2_mean_expr_with_reps.pdf"))
```

```{r}
ggplot(summary, aes(x=group, y=log2(mean_expr+1), fill=group)) +geom_boxplot(outlier.shape=NA, linewidth=0.25) + 
  scale_fill_manual(values=time_cols) + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  facet_wrap(~dar_type) + 
  geom_beeswarm(data=rep_avg, aes(y=log2(mean_expr+1)), cex = 3) +
  coord_cartesian(y=c(0,12)) +
  stat_compare_means(data=rep_avg, comparisons = list(c("day0_ASC","day1_beige"),
                                        c("day1_beige","day3_beige"),
                                        c("day15_white","day0_ASC"),
                                        c("day3_beige","day15_beige"),
                                        c("day15_beige","day0_ASC"), 
                                        c("day15_white","day15_beige")),
                     method="t.test", label="p.format", label.y = seq(from=7, to=11, length.out=6),
                     tip.length =0.01)
ggsave(here("log2_mean_expr_with_reps_&exact_pvals.pdf"))
```



